---
title: "Analysis of US Polution Data"
author: "Isaac Murray"
date: "15 December 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set up libraries.
library(ggplot2)
library(readr)
library(sqldf)
library(RH2)
library(grid)
library(gridExtra)
# Read in dataset.
pol <- read_csv("US_pollution_sample.csv", 
                col_types = cols('Date Local' = col_date(format = "%Y-%m-%d"), 
                                 'NO2 1st Max Value' = col_double()))
# Use to output summary
polsum <- pol

# Dedup dataset.
pol <- unique(pol[!is.na(pol$`CO AQI`) & !is.na(pol$`SO2 AQI`),])

# Add derived variables.
pol$unqref <- paste(as.character(pol$`State Code`),as.character(pol$`Site Num`))
pol$unqsite <- paste(as.character(pol$`County`),as.character(pol$`State`), sep=", ")
pol$unqsitemul <- paste(as.character(pol$`County`),as.character(pol$`State`), sep="\n")
pol$unqsitemul <- gsub(" ", "\n", pol$unqsitemul)
pol$urban <- !(pol$City=="Not in a city")
pol$year <- as.integer(format(pol$`Date Local`,'%Y'))
pol$month <- as.integer(format(pol$`Date Local`,'%m'))
```

# Introduction

This report describes an exploratory data analysis of environmental polution data taken from the United States Environmental Protection Agency (EPA) website. 

# Dataset Description

The original EPA data is contained in multiple files and this has been combined and made available for download. The dataset being analysed is a subset of the downloaded data.  

The following table summarises the dataset variables. Descriptions are copied verbatim from the download site:

| Name              | Type        | Description                                                                   |R Data type |
|:------------------|:------------|:------------------------------------------------------------------------------|:------------|
| State Code        | Categorical |The code allocated by US EPA to each state                                     | int |
| County code       | Categorical |The code of counties in a specific state allocated by US EPA                   | int |
| Site Num          | Categorical | The site number in a specific county allocated by US EPA                      | int |
| Address           | Categorical | State of monitoring site                                                      | chr |
| State             | Categorical | The code allocated by US EPA to each state                                    | int |
| County            | Categorical | County of monitoring site                                                     | chr |
| City              | Categorical | City of the monitoring site                                                   | chr | 
| Date Local        | Categorical | Date of monitoring                                                            | Date |  
| NO2 Units         | Continuous | The units measured for NO2                                                     | num | 
| NO2 Mean          | Continuous | The arithmetic mean of concentration of NO2 within a given day                 | num | 
| NO2 AQI           | Categorical | The calculated air quality index of NO2 within a given day                    | int |
| NO2 1st Max Value | Continuous | The maximum value obtained for NO2 concentration in a given day                | num |
| NO2 1st Max Hour  | Continuous | The hour when the maximum NO2 concentration was recorded in a given day        | num |

# Descriptive statistics and data artefacts
```{r descriptive, echo=FALSE}
#summary(polsum)
```
## Measurement sites and date range of plots  
The dataset includes data collected at seven sites:  
`r paste(unique(pol$unqsite), sep="\n")`

The following plot shows the date ranges of the measurements taken at each site. The gaps indicate missing data. 
```{r missing, echo=FALSE}
ggplot(pol, aes(x=`Date Local`,y=unqsitemul)) + geom_point()
```

## NO2 variables
```{r varNO2, echo=FALSE, message=FALSE, warning=FALSE}
div <- 100
bw <- max(pol$`NO2 Mean`)/div
p1 <- ggplot(data=pol) + aes(x=`NO2 Mean`) + geom_histogram(binwidth=bw) + scale_x_continuous(limits = c(-5, 45))
div <- 100
bw <- max(pol$`NO2 1st Max Value`)/div
p2 <- ggplot(data=pol) + aes(x=`NO2 1st Max Value`) + geom_histogram(binwidth=bw) + scale_x_continuous(limits = c(-5, 45))
bw <- max(pol$`NO2 AQI`)/div
p3 <- ggplot(data=pol) + aes(x=`NO2 AQI`) + geom_bar()# geom_histogram(binwidth=1) 
p4 <- ggplot(data=pol) + aes(x=`NO2 1st Max Hour`) + geom_bar()#geom_histogram(binwidth=1) 
grid.arrange(p1, p2, p3, p4)
```
### Artefacts ###  
#### NO2 1st Max Value  
Notwithstanding a measurement bias towards whole numbers of units, Baltimore, Maryland and Cass, North Dakota sites appear to measure NO2 1st Max Value to a greater precision than the other sites:  
```{r NO2MaxArt, echo=FALSE, message=FALSE, warning=FALSE}
div <- 250
bw <- max(pol$`NO2 1st Max Value`)/div
ggplot(data=pol) + aes(x=`NO2 1st Max Value`) + geom_histogram(binwidth=bw) + scale_x_continuous(limits = c(0,20))+ facet_wrap(~unqsite,scales="free")
```
#### NO2 AQI  
All sites indicate a bias towards the values '8' and '25' for NO2 AQI:  
```{r NO2AQIArt, echo=FALSE, message=FALSE, warning=FALSE}
bw <- max(pol$`NO2 AQI`)/div
ggplot(data=pol) + aes(x=`NO2 AQI`) + geom_bar()+ facet_wrap(~unqsite,scales="free")+ scale_x_continuous(limits = c(6,26))
```

## O3 variables
```{r varO3, echo=FALSE, message=FALSE, warning=FALSE}
div <- 100
bw <- max(pol$`O3 Mean`)/div
p1 <- ggplot(data=pol) + aes(x=`O3 Mean`) + geom_histogram(binwidth=bw)
div <- 200
bw <- max(pol$`O3 1st Max Value`)/div
p2 <- ggplot(data=pol) + aes(x=`O3 1st Max Value`) + geom_histogram(binwidth=bw)
bw <- max(pol$`O3 AQI`)/div
p3 <- ggplot(data=pol) + aes(x=`O3 AQI`) + geom_bar()
p4 <- ggplot(data=pol) + aes(x=`O3 1st Max Hour`) + geom_bar()
grid.arrange(p1, p2, p3, p4)
```

## SO2 variables
```{r varSO2, echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(data=pol) + aes(x=`SO2 Mean`) + geom_histogram(binwidth=0.5)
p2 <- ggplot(data=pol) + aes(x=`SO2 1st Max Value`) + geom_histogram(binwidth=1) + scale_x_continuous(limits = c(-5, 45))
p3 <- ggplot(data=pol) + aes(x=`SO2 AQI`) + geom_bar()
p4 <- ggplot(data=pol) + aes(x=`SO2 1st Max Hour`) + geom_bar() 
grid.arrange(p1, p2, p3, p4)
```
### Artefacts ###  
All sites other than Baltimore, Maryland and arguabley Cass, North Dakota show a disproportionate frequency of zero value measurements.  
```{r artSO2, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=pol) + aes(x=`SO2 Mean`) + geom_histogram(binwidth=0.1)+ facet_wrap(~unqsite,scales="free")
```

## CO variables
```{r varCO, echo=FALSE, message=FALSE, warning=FALSE}
div <- 500
bw <- max(pol$`CO Mean`)/div
p1 <- ggplot(data=pol) + aes(x=`CO Mean`) + geom_histogram(binwidth=bw) #+ scale_x_continuous(limits = c(-.1, 1.2))
p2 <- ggplot(data=pol) + aes(x=`CO 1st Max Value`) + geom_histogram(binwidth=.1) + scale_x_continuous(limits = c(-.1, 1.5))
bw <- max(pol$`CO AQI`)/div
p3 <- ggplot(data=pol) + aes(x=`CO AQI`) + geom_bar()
p4 <- ggplot(data=pol) + aes(x=`CO 1st Max Hour`) + geom_bar()
grid.arrange(p1, p2, p3, p4)
```
### Artefacts ###
Artefacts
All sites other than Baltimore, Maryland a disproportionate frequency of zero measurements.
```{r artCO, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=pol) + aes(x=`SO2 Mean`) + geom_histogram(binwidth=0.1)+ facet_wrap(~unqsite,scales="free")
```



## Line plots of means over time for each polutant
```{r line, echo=FALSE}
ggplot(pol, aes(x=`Date Local`,y=`NO2 Mean`)) + geom_line() + facet_wrap(~unqsite,scales="free")
ggplot(pol, aes(x=`Date Local`,y=`O3 Mean`)) + geom_line() + facet_wrap(~unqsite,scales="free")
ggplot(pol, aes(x=`Date Local`,y=`SO2 Mean`)) + geom_line() + facet_wrap(~unqsite,scales="free")
ggplot(pol, aes(x=`Date Local`,y=`CO Mean`)) + geom_line() + facet_wrap(~unqsite,scales="free")
```

## Box plots of means by month of year for each polutant
```{r box2, echo=FALSE}
ggplot(data=pol) + aes(reorder(month,`NO2 Mean`),`NO2 Mean`) + geom_boxplot() + scale_x_discrete(limits=unique(pol$month)) + facet_wrap(~unqsite,scales="free")
ggplot(data=pol) + aes(reorder(month,`O3 Mean`),`O3 Mean`) + geom_boxplot() + scale_x_discrete(limits=unique(pol$month)) + facet_wrap(~unqsite,scales="free")
ggplot(data=pol) + aes(reorder(month,`SO2 Mean`),`SO2 Mean`) + geom_boxplot() + scale_x_discrete(limits=unique(pol$month)) + facet_wrap(~unqsite,scales="free")
ggplot(data=pol) + aes(reorder(month,`CO Mean`),`CO Mean`) + geom_boxplot() + scale_x_discrete(limits=unique(pol$month)) + facet_wrap(~unqsite,scales="free")
```


You can also embed plots, for example:

```{r pressure, echo=FALSE}
ggplot(pol, aes(x=`Date Local`,y=`NO2 Mean`)) + geom_point(aes(colour = factor(urban),alpha = 1/100)) + facet_wrap(~unqsite,scales="free")
#plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
